---
title: "exploratory_analysis"
author: "Samantha Kotkin, and Sophia Mu"
format: html
---

# Loading Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(knitr)
```

# Data Loading

```{r}
Somerville_data <- read.csv("refined_somerville.csv")
```

# Tables

Correlations between Happiness and Life Satisfaction
```{r}
cor.test(Somerville_data$Somerville.Satisfaction.num, Somerville_data$Happiness.num, method = "pearson")
```
We are 95% confident that there is a moderate (r = .40) and positive correlation between Happiness and Life Satisfaction (p = 2.2e-16), indicating that Happiness is associated with Life Satisfaction. This information helped us to decide to use Happiness as our independent variable in our analysis, as Happiness and Life Satisfaction seem to be interchangeable

Looking at Demographic info
```{r}
countFunction <- function(col){
 Somerville_data |>
    group_by({{col}}) |> 
    summarize(count = n()) 
}
selected <- Somerville_data |> select(Age, Gender, Race.Ethnicity, Household.Income)
#countFunction(c(Age, Gender, Race.Ethnicity))
map(selected, countFunction) |> kable()
#map2(s_data, colnames(s_data), histogramFunction, Somerville_data)
```
Summarized the count of each of the demographic categories that the participants fall into

### Exploratory data analyses
```{r}
#seeing total number of participants
Somerville_data |>
  summarize(count = n())
```

After removing missing variables and answers, there are 471 total participants in the 2025 Somerville Happiness Survey. 

According to the below table, the majority of the participants are White. The next highest amount of participants prefered not to answer. There is a large disparity between Black or African Amerrican and Two or more races respondees and White respondees. This likely reflects on the population of Somerville. 
```{r}
#just seeing number of people in each race because it will def affect analyses
Somerville_data |>
  group_by(Race.Ethnicity) |>
  summarize(raceCount = n()) |>
  kable()
```


```{r}
#function pt 2
findingMean <- function(data){
 data |>
  group_by(Race.Ethnicity) |>
  summarize(across(c(Education.Quality.Satisfaction.num, Housing.Condition.Satisfaction.num, Feel.Part.Community.num, Somerville.Satisfaction.num, Happiness.num), mean))
}
```

```{r}
findingMean(Somerville_data) |> kable()
```
This table shows us the average happiness and satisfaction scores across different survey variables for each race. Participants who did not identify their race/ethnicity were on average the least happy with their Education Quality, followed by White participants. Participants with two or more races/ethnicity were on average the least satisfied with their housing. Participants who did not identify their race/ethnicity on average felt the least a part of their community.

## Histogram function
```{r}
s_data <- Somerville_data |>
  select(Education.Quality.Satisfaction.num, Housing.Condition.Satisfaction.num, Feel.Part.Community.num, Somerville.Satisfaction.num, Happiness.num)
```

```{r}
histogramFunction <- function(col, label, data){
  data |> 
  ggplot(aes(x = {{col}})) + 
  geom_histogram() +
  labs(x = label)
}
```

```{r}
map2(s_data, colnames(s_data), histogramFunction, Somerville_data)
```

Histogram showing the range of Satisfaction of all participants for each variable. Across all variables, participants were the least satisfied with the quality of each given variable, with ratings in the 4s and 5s. Because of this trend, most of these histograms are skewed left. Education quality satisfaction seems to be more normally distributed, with the majority of the participants rating it a 3. 
